---
title: "Creating Patient Identifiers and Time-Based Episodes with epidm"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Creating Patient Identifiers and Time-Based Episodes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(epidm)
library(data.table)

```


# Introduction

The **epidm** R package provides utilities to support patient identity
resolution and clinical episode construction.

This vignette introduces two key functions:

1. **`uk_patient_id()`**  
   Generates unique patient identifiers using patient
   identifiers with multiple staged rules.

2. **`group_time()`**  
   Groups time intervals into clinically meaningful
   episodes (static or rolling windows).

Both functions are designed for **healthcare data**, for example SGSS, HES/SUS data
and ECDS data where repeated records across systems must be linked or aggregated.

# 1. `uk_patient_id()` function

## Input Requirements

A dataset should ideally contain:

- NHS number
- Hospital number
- Date of birth
- Sex
- Forename
- Surname
- Postcode

You must also supply a **list mapping your column names** using the `id = list(...)` argument.

## Example usage

```{r ukpid-example, echo=TRUE}
# 1) Create example data (shown so readers can reproduce)
id_test <-
  data.frame(
    stringsAsFactors = FALSE,
    record_id = c(1L,2L,3L,4L,
                  5L,6L,7L,8L,9L,10L,11L,12L,13L,14L,15L,
                  16L,17L,18L,19L,20L,21L,22L,23L,24L),
    nhs_number = c(9435754422,
                   9435754422,NA,9435754422,5555555555,NA,
                   9435773982,NA,9999999999,NA,9435773982,NA,
                   9435802508,9435802508,NA,NA,9435802508,9435802508,NA,
                   3333333333,NA,9999999999,9435817777,
                   9435817777),
    local_patient_identifier = c(NA,"IG12067",
                                 NA,NA,"IG12067","IG12067","KR2535","KR2535",
                                 "KR2535",NA,NA,NA,"UK8734","UK8734",NA,NA,
                                 "UK8734","UK8734",NA,NA,"JH45204",
                                 "HS45202","HS45202","JH45204"),
    patient_birth_date = c("1993-07-16",
                           "1993-07-16","1993-07-16","1993-07-16",
                           "1993-07-16",NA,"1967-02-10",NA,"1967-02-10",NA,NA,
                           "1967-02-10",NA,NA,"1952-10-22","1952-10-22",
                           "1952-10-22",NA,"1947-09-14","1947-09-14",
                           "1947-09-14","1947-09-14","1947-09-14",
                           "1947-09-14"),
    sex = c("Male","Male",
            "Male","Male",NA,"Male","Female","Female",
            "Female","Female","Female","Female","Male",
            "Male","Male","Male","Male","Male","Male",
            "Male","Male","Male",NA,"Male"),
    forename = c(NA,"DENNIS",
                 NA,NA,"DENNIS",NA,"ELLIE","ELLIE",NA,
                 "ELLIE","ELLIE","ELLIE","IAN","IAN","MALCOLM",
                 "IAN","IAN",NA,"GRANT","ALAN","ALAN","ALAN",
                 "GRANT","ALAN"),
    surname = c(NA,"NEDRY",
                "NEDRY",NA,"NEDRY","NEDRY","SATTLER","SATTLER",
                NA,"SATTLER","SATTLER","SATTLER","M",NA,
                "IAN","MALCOLM","MALCOLM",NA,"ALAN","GRANT",
                "GRANT","GRANT","ALAN","GRANT"),
    postcode = c("HA4 0FF",
                 "HA4 0FF","HA4 0FF",NA,"HA4 0FF","HA4 0FF",
                 "L3 1DZ","L3 1DZ","L3 1DZ","L3 1DZ",NA,"L3 1DZ",
                 "BN14 9EP",NA,"BN14 9EP",NA,NA,NA,"CW6 9TX",
                 "CW6 9TX",NA,NA,NA,NA),
    specimen_date = c("2024-08-14",
                      "2023-02-03","2023-02-07","2023-02-04",
                      "2023-02-09","2024-08-14","2021-03-28","2021-03-28",
                      "2021-03-28","2021-03-28","2021-03-28",
                      "2021-03-28","2024-07-06","2024-07-06","2024-07-06",
                      "2023-10-31","2023-10-31","2023-10-31",
                      "2022-01-23","2022-01-24","2022-01-25","2022-01-26",
                      "2022-01-27","2022-01-28")
  )

# 2) Run uk_patient_id()
result_id <- uk_patient_id(
  id_test,
  id = list(
    nhs_number    = 'nhs_number',
    hospital_number = 'local_patient_identifier',
    date_of_birth = 'patient_birth_date',
    sex_mfu       = 'sex',
    forename      = 'forename',
    surname       = 'surname',
    postcode      = 'postcode'
  ),
  .useStages   = 1:11,      # optional
  .keepStages  = FALSE,     # optional
  .keepValidNHS = FALSE     # optional
)

# 3) Show a compact preview
print(head(result_id), row.names = FALSE)
```

# 2. `group_time()` function

## Input Requirements

A dataset should ideally contain:

- One or more grouping variables (e.g., patient ID, organism, specimen type)
- A start date column (date_start)
- For intervals: an end date column (date_end)
- For events: a window (integer, days) and window_type ("static" or "rolling")

You must also supply the grouping and date arguments via function parameters:

- group_vars (character vector of column names)
- date_start
- plus either date_end (for intervals) or window + window_type (for events).
- Optional: indx_varname, min_varname, max_varname to control output column names

## Purpose

`group_time()` aggregates:

- **Intervals** (start + end dates)  
  e.g., hospital spells (HES/SUS)

- **Events** (single date)  
  e.g., microbiology specimen dates

Two episode rules are supported:

| Type | Meaning |
|------|---------|
| **Static window** | Based on the first event only; all events within X days of the **first** event belong to the same episode. |
| **Rolling window** | The window resets with each event; a new event inside X days of the **previous** event extends the episode. |

```{r grouptime-example, echo=TRUE}
# Events example (14‑day static window):
#1) Create example data

episode_test <- structure(
  list(
    pat_id = c(1L, 1L, 1L, 1L, 2L, 2L, 2L, 1L, 1L, 1L, 1L, 2L, 2L, 2L),
    species = c(rep("E. coli", 7), rep("K. pneumonia", 7)),
    spec_type = c(rep("Blood", 7), rep("Blood", 4), rep("Sputum", 3)),
    sp_date = structure(
      c(
        18262,
        18263,
        18281,
        18282,
        18262,
        18263,
        18281,
        18265,
        18270,
        18281,
        18283,
        18259,
        18260,
        18281
      ),
      class = "Date"
    )
  ),
  row.names = c(NA, -14L),
  class = "data.frame"
)

# 2) Run group_time() for events using a 14-day static window
ep_static <- group_time(
  x = episode_test,
  date_start = 'sp_date',
  window = 14,
  window_type = 'static',
  group_vars = c('pat_id', 'species', 'spec_type'),
  indx_varname = 'static_indx',   # optional
  min_varname = 'min_date',
  # optional (defaults)
  max_varname = 'max_date'      # optional (defaults)
)
# 3) Show a compact preview
print(head(ep_static), row.names = FALSE)

# Intervals example (start + end dates):1) Create example interval data (start + end dates)
#1) Create example data
spell_test <- data.frame(
  id = c(rep(99, 6), rep(88, 4), rep(3, 3)),
  provider = c("YXZ", rep("ZXY", 5), rep("XYZ", 4), rep("YZX", 3)),
  spell_start = as.Date(
    c(
      "2020-03-01",
      "2020-07-07",
      "2020-02-08",
      "2020-04-28",
      "2020-03-15",
      "2020-07-01",
      "2020-01-01",
      "2020-01-12",
      "2019-12-25",
      "2020-03-28",
      "2020-01-01",
      NA,
      NA
    )
  ),
  spell_end = as.Date(
    c(
      "2020-03-10",
      "2020-07-26",
      "2020-05-22",
      "2020-04-30",
      "2020-05-20",
      "2020-07-08",
      "2020-01-23",
      "2020-03-30",
      "2020-01-02",
      "2020-04-20",
      "2020-01-01",
      NA,
      NA
    )
  )
)

# 2) Run group_time() for intervals (start + end dates)
spell_episodes <- group_time(
  x = spell_test,
  date_start = 'spell_start',
  date_end = 'spell_end',
  group_vars = c('id', 'provider'),
  indx_varname = 'spell_id',
  # optional
  min_varname = 'spell_min_date',
  # optional
  max_varname = 'spell_max_date'        # optional
)

# 3) Show a compact preview
print(head(spell_episodes), row.names = FALSE)
```

